Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent

### 2023-05-29 manuscript/main.tex
more complete formula

\usepackage{algorithm2e}
\usepackage[usenames,dvipsnames]{xcolor}
\SetAlgoLined
\SetKwProg{for}{For}{:}{end}
\SetKwProg{pgsgd}{PG-SGD}{:}{end}
\SetKwProg{each}{For each}{:}{end}
\usepackage{bm}

@@

\paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is theoretically applicable in any number of dimensions.
In this work, we describe the results in 1D and 2D.
In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
\begin{itemize}
	\item The first node Xi of a pair is a uniform path step pick from all nodes.
	\item The second node Xj of a pair is sampled from the same path following a Zipfian distribution. 
	\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes. The magnitude r of the update depends on the current learning rate of the SGD.
\end{itemize}
%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread. We can have several worker threads but only one checker thread!

\begin{algorithm}
	\caption{Pseudocode of the PG-SGD algorithm. Explain $\eta$. Explain $PathIndex$ in the text. Explain $ZipfZetas$ in the text. Explain $RandomLayout$ in the text. Explain annealing schedule in the text or rephrase this wording.}
	\label{alg:pg_sgd}
	\pgsgd{\textbf{(}V\textbf{)}}{
		\textbf{input:} variation graph $V = (N, E, P)$ \\
		%https://tex.stackexchange.com/questions/22643/how-to-write-letters-in-bold-in-the-math-mode
		\textbf{output:} $k$-dimensional layout \boldsymbol{$L$} with $n$ nodes \\
		\boldsymbol{$P$} $\leftarrow PathIndex(G)$ \\ 
		\boldsymbol{$Z$} $\leftarrow ZipfZetas(G,P)$ \\ %@Andrea is this correct like this?
		\boldsymbol{$L$} $\leftarrow RandomLayout(n, k)$ \tcp{only 2D PG-SGD has a random layout initialization, 1D starts with the current node order}
		% atomic positions initialization?
		% TODO for somplicity reasons I would only describe the 2D one here
		\for{$\eta$ $in$ $annealing$ $schedule$}{
			\each{$term$ $update$} {
				random path step from all steps \\
				if cooling $||$ flip \\
				uniform step across the path \\
				else \\
				forward \\
				use Z to generate max jump length \\
				newsteprank = cursteprank + sampledzipfvalue \\
				backward \\
				use Z to generate jump length \\
				newsteprank = cursteprank + sampledzipfvalue \\
				get path positions via path index \\	
				$nuc\_dist=abs(pos\_in\_path\_a - pos\_in\_path\_b)$ \\
				$term\_weight = 1.0 / term\_dist$ \\
				$w\_ij = term\_weight$ \\
				$mu = eta.load() * w_{ij}$ \\
				$dx = X[i].load() - X[j].load()$ \tcp{calculate layout distance }
				$mag = std::abs(dx)$ \\
				$Delta = mu * (mag - d_ij) / 2$ \\
				$ Delta_abs = std::abs(Delta)$ \\
				\tcp{potentially store new Delta\_max}
				$r = Delta / mag$ \\
				$ r_x = r * dx$ \\
				\tcp{new X = old X - r\_x}
				\tcp{new Y = old Y - r\_Y, 2D only}
			}
		}
	}
\end{algorithm}

>>>>

\usepackage{algorithm2e}
\usepackage[usenames,dvipsnames]{xcolor}
\SetAlgoLined
\usepackage{bm}

@@

\paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is theoretically applicable in any number of dimensions.
In this work, we describe the results in 1D and 2D.
In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
This allows the PG-SGD to be applied manuscriptto gigabase-scale pangenome graphs. \\
Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
\begin{itemize}
	\item The first node Xi of a pair is a uniform path step pick from all nodes.
	\item The second node Xj of a pair is sampled from the same path following a Zipfian distribution. 
	\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes. The magnitude r of the update depends on the current learning rate of the SGD.
\end{itemize}
%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread. We can have several worker threads but only one checker thread!

\input{pg-sgd_pseudocode}
